Asymmetric relationships and their effects on coexistence: Supporting information

### 2023-02-16 paper/SupMat.Rnw
finished Methods and accompanying SI sections, at least for now
fixed an error spotted by Jasmin
fixed typo in one of the sup mat formulas that I noticed while doing something else. it's not consequential.
made some aspects clearer of how efficient computations are done

We next decompose $\olr_{j \bs i}$ for the beta fecundities lottery model [see (\ref{eq:decomp_overall_rjbsi})]. We have 
\begin{equation}
E_j = B_j = \eta_j \beta_j
\end{equation}
\begin{equation}
C_{j \bs i} = \frac{B_j}{\delta} = \frac{\eta_j \beta_j}{\delta}
\end{equation}
\begin{equation}
\overline{E}_j = \frac{\eta_j}{2}
\end{equation}
\begin{equation}
\overline{C}_{j \bs i} = \frac{\eta_j}{2\delta}
\end{equation}
\begin{equation}
\overline{B}_j = \frac{\eta_j}{2}
\end{equation}
\begin{align}
\olr_{j \bs i} &= \E\left[ \ln\left( 1-\delta+\frac{E_j}{C_{j \bs i}} \right) \right] \\
&= \E\left[ \ln\left( 1-\delta+\frac{B_j}{B_j} \right) \right] \\
&= 0 \label{eq:119}
\end{align}
\begin{equation}
\epsilon_j^0 = \ln \left( 1-\delta+\delta\frac{\eta_j /2}{\eta_j /2} \right) = 0 \label{eq:120}
\end{equation}
\begin{align}
\olep_j^E &= \E\left[ \ln\left( 1-\delta+\delta\frac{2\eta_j \beta_j}{\eta_j} \right) \right] - \epsilon_j^0 \\
&= \E\left[ \ln\left( 1-\delta+2 \delta \beta_j \right) \right] \label{eq:122}
\end{align}
\begin{align}
\olep_j^C &= \E\left[ \ln\left( 1-\delta+\delta\frac{\eta_j}{2\eta_j\beta_j} \right) \right]-\epsilon_j^0 \\
&= \E\left[ \ln\left( 1-\delta+\frac{\delta}{2\beta_j} \right) \right] \label{eq:124}
\end{align}
\begin{align}
\olep_j^{EC} &= \E\left[  \ln\left( 1-\delta+\delta\frac{B_j}{B_j} \right)\right]-\olep_j^E-\olep_j^C-\epsilon_j^0 \\
&= -\olep_j^E - \olep_j^C
\end{align}
\begin{align}
\olep_j^{(E\shp C)} &= \E\left[ \ln\left( 1-\delta+\delta\frac{B_j}{B_j^\shp} \right) \right]-\olep_j^E-\olep_j^C-\epsilon_j^0 \\
&= \E\left[ \ln\left( 1-\delta+\delta\frac{\beta_j}{\beta_j^\shp} \right) \right]-\olep_j^E-\olep_j^C, \label{eq:128}
\end{align}
where $(\beta_j,\beta_j^\shp)=(F_\beta^{-1} \circ \varphi (b_j),F_\beta^{-1} \circ \varphi (b_j^\shp))$ and
$b_j^\shp$ is distributed in the same way as $b_j$ but is independent of it. Continuing,
\begin{align}
\olep_j^{(EC)} &= \E\left[ \ln\left(1-\delta+\delta \frac{B_j}{B_j} \right)  \right]-
\E\left[ \ln\left(1-\delta+\delta\frac{B_j}{B_j^\shp}  \right)  \right] \\
&= -\E\left[\ln\left( 1-\delta+\delta\frac{\beta_j}{\beta_j^\shp} \right)  \right].
\end{align}
Continuing,
\begin{align}
\olep_j^{[EC]} &= \E\left[ \ln\left(1-\delta+\delta \frac{B_j}{B_j} \right)  \right]-
\E\left[ \ln\left(1-\delta+\delta\frac{B_j^\nat}{B_j^\nat}  \right)  \right] \\
&= 0.\label{eq:132}
\end{align}
We showed already (SI section \ref{SIsect:TheoryForLotteryBoth} and 
\ref{SIsect:TheoryForLotteryLogNorm}) that this is $0$ because $E_j$ and $C_{j \bs i}$
are linearly related to each other. Continuing,
\begin{align}
\olep_j^{[E \nat C]} &= \olep_j^{(EC)}-\olep_j^{[EC]} \\
&= -\E\left[ \ln\left( 1-\delta+\delta\frac{\beta_j}{\beta_j^\shp}  \right)  \right].\label{eq:134}
\end{align}
This completes the calculation of all the terms of the decomposition of $\olr_{j \bs i}$
for the beta fecundities lottery model.

@@

First, for a large integer $M$ (we used $M=\Sexpr{M}$),
carry out the following steps once:
\begin{enumerate}
\item Generate $M$ independent points $(\beta_i^{(k)},\beta_j^{(k)})=(F_\beta^{-1} \circ \varphi (b_i^{(k)}),F_\beta^{-1} \circ \varphi (b_j^{(k)}))$,
$k=1,\ldots,M$, where $(b_i^{(k)},b_j^{(k)})$ are left- or right-tail associated points generated as in SI section \ref{sect:noise},
depending on whether one is currently considering the left- or right-tail associated case.
\item Generate $M$ independent points $(\beta_i^{\shp,(k)},\beta_j^{\shp,(k)})=(F_\beta^{-1} \circ \varphi (b_i^{\shp,(k)}),
F_\beta^{-1} \circ \varphi (b_j^{\shp,(k)}))$, $k=1,\ldots,M$, where $b_i^{\shp,(k)}$ and $b_j^{\shp,(k)}$ are standard normal and 
independent.
\item Generate $M$ independent points 
$(\beta_i^{\nat,(k)},\beta_j^{\nat,(k)})=(F_\beta^{-1} \circ \varphi (b_i^{\nat,(k)}),
F_\beta^{-1} \circ \varphi (b_j^{\nat,(k)}))$, $k=1,\ldots,M$, where the points $(b_i^{\nat,(k)},b_j^{\nat,(k)})$
come from a bivariate normal distribution with standard normal marginals and Pearson correlation the same
as $(b_i^{(k)},b_j^{(k)})$ (see SI section \ref{sect:noise}). 
\end{enumerate}
These data are stored and used repeatedly below.

@@

Next, $q_{ij}$ can be taken (SI section \ref{sect:noise}) to be, alternatively, either $1$ or the quantity displayed in 
(\ref{eq:137}). Hence the terms of the decomposition (\ref{eq:decomp_overall_delta}) can straightforwardly be
estimated using our estimates elaborated above, as follows. There are two versions of each estimator, corresponding
to the alternative values of $q_{ij}$. 
%***DAN: stopped here, 2023 02 03. You can continue from p. 8 of your hand-written version of this

>>>>

We next decompose $\olr_{j \bs i}$ for the beta fecundities lottery model [see (\ref{eq:decomp_overall_rjbsi})]. We have 
\begin{equation}
E_j = B_j = \eta_j \beta_j
\end{equation}
\begin{equation}
C_{j \bs i} = \frac{B_j}{\delta} = \frac{\eta_j \beta_j}{\delta}
\end{equation}
\begin{equation}
\overline{E}_j = \frac{\eta_j}{2}
\end{equation}
\begin{equation}
\overline{C}_{j \bs i} = \frac{\eta_j}{2\delta}
\end{equation}
\begin{equation}
\overline{B}_j = \frac{\eta_j}{2}
\end{equation}
\begin{align}
\olr_{j \bs i} &= \E\left[ \ln\left( 1-\delta+\frac{E_j}{C_{j \bs i}} \right) \right] \\
&= \E\left[ \ln\left( 1-\delta+\delta\frac{B_j}{B_j} \right) \right] \\
&= 0 \label{eq:119}
\end{align}
\begin{equation}
\epsilon_j^0 = \ln \left( 1-\delta+\delta\frac{\eta_j /2}{\eta_j /2} \right) = 0 \label{eq:120}
\end{equation}
\begin{align}
\olep_j^E &= \E\left[ \ln\left( 1-\delta+\delta\frac{2\eta_j \beta_j}{\eta_j} \right) \right] - \epsilon_j^0 \\
&= \E\left[ \ln\left( 1-\delta+2 \delta \beta_j \right) \right] \label{eq:122}
\end{align}
\begin{align}
\olep_j^C &= \E\left[ \ln\left( 1-\delta+\delta\frac{\eta_j}{2\eta_j\beta_j} \right) \right]-\epsilon_j^0 \\
&= \E\left[ \ln\left( 1-\delta+\frac{\delta}{2\beta_j} \right) \right] \label{eq:124}
\end{align}
\begin{align}
\olep_j^{EC} &= \E\left[  \ln\left( 1-\delta+\delta\frac{B_j}{B_j} \right)\right]-\olep_j^E-\olep_j^C-\epsilon_j^0 \\
&= -\olep_j^E - \olep_j^C
\end{align}
\begin{align}
\olep_j^{(E\shp C)} &= \E\left[ \ln\left( 1-\delta+\delta\frac{B_j}{B_j^\shp} \right) \right]-\olep_j^E-\olep_j^C-\epsilon_j^0 \\
&= \E\left[ \ln\left( 1-\delta+\delta\frac{\beta_j}{\beta_j^\shp} \right) \right]-\olep_j^E-\olep_j^C, \label{eq:128}
\end{align}
where $(\beta_j,\beta_j^\shp)=(F_\beta^{-1} \circ \varphi (b_j),F_\beta^{-1} \circ \varphi (b_j^\shp))$ and
$b_j^\shp$ is distributed in the same way as $b_j$ but is independent of it. Continuing,
\begin{align}
\olep_j^{(EC)} &= \E\left[ \ln\left(1-\delta+\delta \frac{B_j}{B_j} \right)  \right]-
\E\left[ \ln\left(1-\delta+\delta\frac{B_j}{B_j^\shp}  \right)  \right] \\
&= -\E\left[\ln\left( 1-\delta+\delta\frac{\beta_j}{\beta_j^\shp} \right)  \right].
\end{align}
Continuing,
\begin{align}
\olep_j^{[EC]} &= \E\left[ \ln\left(1-\delta+\delta \frac{B_j}{B_j} \right)  \right]-
\E\left[ \ln\left(1-\delta+\delta\frac{B_j^\nat}{B_j^\nat}  \right)  \right] \\
&= 0.\label{eq:132}
\end{align}
We showed already (SI section \ref{SIsect:TheoryForLotteryBoth} and 
\ref{SIsect:TheoryForLotteryLogNorm}) that this is $0$ because $E_j$ and $C_{j \bs i}$
are linearly related to each other. Continuing,
\begin{align}
\olep_j^{[E \nat C]} &= \olep_j^{(EC)}-\olep_j^{[EC]} \\
&= -\E\left[ \ln\left( 1-\delta+\delta\frac{\beta_j}{\beta_j^\shp}  \right)  \right].\label{eq:134}
\end{align}
This completes the calculation of all the terms of the decomposition of $\olr_{j \bs i}$
for the beta fecundities lottery model.

@@

First, for a large integer $M$ (we used $M=\Sexpr{M}$),
carry out the following steps once:
\begin{enumerate}
\item Generate $M$ independent points $(\beta_i^{(k)},\beta_j^{(k)})=(F_\beta^{-1} \circ \varphi (b_i^{(k)}),F_\beta^{-1} \circ \varphi (b_j^{(k)}))$,
$k=1,\ldots,M$, where $(b_i^{(k)},b_j^{(k)})$ are left- or right-tail associated points generated as in SI section \ref{sect:noise},
depending on whether one is currently considering the left- or right-tail associated case.
\item Generate $M$ independent points $(\beta_i^{\shp,(k)},\beta_j^{\shp,(k)})=(F_\beta^{-1} \circ \varphi (b_i^{\shp,(k)}),
F_\beta^{-1} \circ \varphi (b_j^{\shp,(k)}))$, $k=1,\ldots,M$, where $b_i^{\shp,(k)}$ and $b_j^{\shp,(k)}$ are standard normal and 
independent. Make these points also independent of the points $(\beta_i^{(k)},\beta_j^{(k)})$ generated above.
\item Generate $M$ independent points 
$(\beta_i^{\nat,(k)},\beta_j^{\nat,(k)})=(F_\beta^{-1} \circ \varphi (b_i^{\nat,(k)}),
F_\beta^{-1} \circ \varphi (b_j^{\nat,(k)}))$, $k=1,\ldots,M$, where the points $(b_i^{\nat,(k)},b_j^{\nat,(k)})$
come from a bivariate normal distribution with standard normal marginals and Pearson correlation the same
as $(b_i^{(k)},b_j^{(k)})$ (see SI section \ref{sect:noise}). Make these points also independent of both sets of 
points generated above.
\end{enumerate}
These data are stored and used repeatedly below.

@@

Next, $q_{ij}$ can be taken (SI section \ref{sect:noise}) to be, alternatively, either $1$ or the quantity displayed in 
(\ref{eq:137}). Hence the terms of the decomposition (\ref{eq:decomp_overall_delta}) can straightforwardly be
estimated using our estimates elaborated above, as follows. There are two versions of each estimator, corresponding
to the alternative values of $q_{ij}$. 
\begin{enumerate}
%baseline
\item Compute $\Delta_i^0 = \epsilon_i^0 - q_{ij} \epsilon_j^0$, which is exact. Since $\epsilon_j^0=0$, 
we have $\Delta_i^0 = \epsilon_i^0$.
%E
\item Estimate $\Delta_i^E$ as $\widehat{\Delta_i^E}=\widehat{\olep_i^E}-q_{ij} \widehat{\olep_j^E}=
\mean_k[\ln(1-\delta+2\delta\frac{\eta_i}{\eta_j}\beta_i^{\shp,(k)})]-\epsilon_i^0-
q_{ij} \mean_k[\ln(1-\delta+2\delta\beta_j^{(k)})]$, which has standard error 
$\se(\widehat{\Delta_i^E})=\sqrt{s_1^2+s_2^2}$, where $s_1=\sd_k[\ln(1-\delta+2\delta\frac{\eta_i}{\eta_j}\beta_i^{\shp,(k)})]/\sqrt{M}$
and $s_2=q_{ij}\sd_k[\ln(1-\delta+2\delta\beta_j^{(k)})]/\sqrt{M}$.
%C
\item Estimate $\Delta_i^C$ as $\widehat{\Delta_i^C}=\widehat{\olep_i^C}-q_{ij}\widehat{\olep_j^C}=
\mean_k[\ln(1-\delta+\frac{\delta}{2}\frac{\eta_i}{\eta_j}\frac{1}{\beta_j^{\shp,(k)}})]-\epsilon_i^0-q_{ij} 
\mean_k[\ln(1-\delta+\frac{\delta}{2}\frac{1}{\beta_j^{\shp,(k)}})]$, which has standard error
$\se(\widehat{\Delta_i^C})=\sd_k[\ln(1-\delta+\frac{\delta}{2}\frac{\eta_i}{\eta_j}\frac{1}{\beta_j^{\shp,(k)}})-
q_{ij}\ln(1-\delta+\frac{\delta}{2}\frac{1}{\beta_j^{\shp,(k)}})]/\sqrt{M}$.
%E#C
\item Estimate $\Delta_i^{(E\shp C)}$ as $\widehat{\Delta_i^{(E\shp C)}}=\widehat{\olep_i^{(E\shp C)}}-q_{ij}
\widehat{\olep_j^{(E\shp C)}}$. This has standard error 
$\se(\widehat{\Delta_i^{(E\shp C)}})=\sd_k[\ln(1-\delta+\delta\frac{\eta_i}{\eta_j}\frac{\beta_i^{\shp,(k)}}{\beta_j^{\shp,(k)}})-
\ln(1-\delta+2\delta\frac{\eta_i}{\eta_j}\beta_i^{\shp,(k)})-\ln(1-\delta+\frac{\delta}{2}\frac{\eta_1}{\eta_2}\frac{1}{\beta_j^{\shp,(k)}})-
q_{ij}\ln(1-\delta+\delta\frac{\beta_j^{(k)}}{\beta_j^{\shp,(k)}})+q_{ij}\ln(1-\delta+2\delta\beta_j^{(k)})+
q_{ij}\ln(1-\delta+\frac{\delta}{2}\frac{1}{\beta_j^{\shp,(k)}})]/\sqrt{M}$.
%E||C
\item Estimate $\Delta_i^{[E\nat C]}$ as $\widehat{\Delta_i^{[E\nat C]}}=\widehat{\olep_i^{[E\nat C]}}-
q_{ij}\widehat{\olep_j^{[E\nat C]}}=
\mean_k[\ln(1-\delta+\delta\frac{\eta_i}{\eta_j}\frac{\beta_i^{\nat,(k)}}{\beta_j^{\nat,(k)}})]-
\mean_k[\ln(1-\delta+\delta\frac{\eta_i}{\eta_j}\frac{\beta_i^{\shp,(k)}}{\beta_j^{\shp,(k)}})]+
q_{ij}\mean_k[\ln(1-\delta+\delta\frac{\beta_j^{(k)}}{\beta_j^{\shp,(k)}})]$, 
which has standard error
$\se(\widehat{\Delta_i^{[E\nat C]}})=\sd_k[\ln(1-\delta+\delta\frac{\eta_i}{\eta_j}\frac{\beta_i^{\nat,(k)}}{\beta_j^{\nat,(k)}})-
\ln(1-\delta+\delta\frac{\eta_i}{\eta_j}\frac{\beta_i^{\shp,(k)}}{\beta_j^{\shp,(k)}})+
q_{ij}\ln(1-\delta+\delta\frac{\beta_j^{(k)}}{\beta_j^{\shp,(k)}})]/\sqrt{M}$.
%[EC]
\item For $\Delta_i^{[EC]}$, we recall that $\olep_j^{[EC]}=0$, so $\Delta_i^{[EC]}=\olep_i^{[EC]}$.
So we use $\widehat{\olep_i^{[EC]}}$ for $\Delta_i^{[EC]}$.
%r
\item Finally, we can estimate $\olr_{i \bs i}-q_{ij} \olr_{j \bs i}$ simply as $\widehat{\olr_{i \bs i}}$,
since $\olr_{j \bs i}=0$.
\end{enumerate}
